---
title: "Interfaces for Bioinformatics Containers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interfaces for Bioinformatics Containers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

```{r setup}
library(S7)
library(s7contract)
```

Bioinformatics packages often exchange rich containers rather than plain
matrices.  Bioconductor's `SummarizedExperiment` class, for example, coordinates
assays, feature metadata, sample metadata, names, validity rules, and many
specialized methods.  This vignette does **not** reimplement that class.  It
asks a narrower design question: can interfaces describe a small slice of such a
container so downstream code can depend on behavior rather than one concrete
class?

## A toy assay container

We make a tiny S7 class with assays, row metadata, and column metadata.  The
validator enforces the minimum shape invariant needed by the examples.

```{r toy-container}
MiniSummarizedExperiment <- new_class(
  "MiniSummarizedExperiment",
  properties = list(
    assays = class_list,
    row_data = class_data.frame,
    col_data = class_data.frame
  ),
  validator = function(self) {
    if (length(self@assays) == 0) {
      return("@assays must contain at least one matrix")
    }

    dims <- lapply(self@assays, dim)
    if (any(vapply(dims, is.null, logical(1)))) {
      return("every assay must be matrix-like")
    }

    first_dim <- dims[[1]]
    same_dim <- vapply(dims, identical, logical(1), first_dim)
    if (!all(same_dim)) {
      return("all assays must have the same dimensions")
    }

    if (nrow(self@row_data) != first_dim[[1]]) {
      return("@row_data must have one row per assay feature")
    }
    if (nrow(self@col_data) != first_dim[[2]]) {
      return("@col_data must have one row per assay sample")
    }
  }
)

counts <- matrix(
  c(10, 0, 3, 4, 12, 8),
  nrow = 3,
  dimnames = list(c("geneA", "geneB", "geneC"), c("sample1", "sample2"))
)

mini <- MiniSummarizedExperiment(
  assays = list(counts = counts, logcounts = log1p(counts)),
  row_data = data.frame(gc = c(0.42, 0.51, 0.37), row.names = rownames(counts)),
  col_data = data.frame(condition = c("control", "treated"), row.names = colnames(counts))
)
```

## An interface for the behavior a consumer needs

A differential-expression helper might not care about the concrete class.  It
might only need assay names, feature names, sample names, and a way to retrieve
an assay matrix.

```{r assay-interface}
assay_names <- new_generic("assay_names", "x")
feature_names <- new_generic("feature_names", "x")
sample_names <- new_generic("sample_names", "x")
assay_matrix <- new_generic("assay_matrix", "x")

AssayContainer <- new_interface(
  "AssayContainer",
  methods = list(
    assay_names = assay_names,
    feature_names = feature_names,
    sample_names = sample_names,
    assay_matrix = assay_matrix
  )
)

method(assay_names, MiniSummarizedExperiment) <- function(x) names(x@assays)
method(feature_names, MiniSummarizedExperiment) <- function(x) rownames(x@assays[[1]])
method(sample_names, MiniSummarizedExperiment) <- function(x) colnames(x@assays[[1]])
method(assay_matrix, MiniSummarizedExperiment) <- function(x, name = assay_names(x)[[1]]) {
  x@assays[[name]]
}

implements(mini, AssayContainer)
assay_names(mini)
sample_names(mini)
assay_matrix(mini, "counts")[, "sample1"]
```

A consumer can now assert the behavior rather than a concrete class name.

```{r assay-consumer}
library_size <- function(x, assay = "counts") {
  assert_implements(x, AssayContainer)
  mat <- assay_matrix(x, assay)
  colSums(mat)
}

library_size(mini)
```

This is the useful part of the idea: an interface can describe a *small view* of
an object.  Another package could satisfy the same interface with an HDF5-backed
matrix container, a remote query result, or a test double, as long as the same
runtime methods exist.

## A nominal trait for explicit adapters

The trait layer is useful when structural compatibility is not enough.  Here the
implementation records extra metadata about the orientation of the assay.

```{r assay-trait, message = FALSE, warning = FALSE}
ExperimentLike <- new_trait(
  "ExperimentLike",
  methods = list(
    assay_names = trait_method(assay_names),
    feature_names = trait_method(feature_names),
    sample_names = trait_method(sample_names),
    assay_matrix = trait_method(assay_matrix)
  ),
  assoc_consts = c("ASSAY_ORIENTATION")
)

impl_trait(
  ExperimentLike,
  MiniSummarizedExperiment,
  methods = list(
    assay_names = function(x) names(x@assays),
    feature_names = function(x) rownames(x@assays[[1]]),
    sample_names = function(x) colnames(x@assays[[1]]),
    assay_matrix = function(x, name = assay_names(x)[[1]]) x@assays[[name]]
  ),
  assoc_consts = list(ASSAY_ORIENTATION = "features_by_samples"),
  replace = TRUE
)

has_trait(mini, ExperimentLike)
trait_assoc_const(ExperimentLike, mini, "ASSAY_ORIENTATION")
```

## Where the mimicry breaks

Interfaces are not a substitute for `SummarizedExperiment` itself.

* They do not reproduce Bioconductor's S4 method ecosystem, validity contracts,
  delayed operations, genomic ranges integration, or metadata conventions.
* They do not enforce semantic laws such as row metadata staying synchronized
  after subsetting unless the class and its methods implement those laws.
* They do not make assay data fast.  Performance still belongs in the concrete
  matrix/container implementation.
* They cannot prove at compile time that a downstream analysis is safe.

## Nonsense ideas and sharper alternatives

| Nonsense idea | Critique | Sharper alternative |
|---|---|---|
| "Let's rebuild `SummarizedExperiment` with interfaces." | The existing class is a mature interoperability standard.  Rebuilding it would fragment users and lose a large method ecosystem. | Use interfaces for small adapters or tests while respecting established container classes. |
| "An interface should guarantee biological correctness." | A method list cannot prove that batches, features, genome builds, and assay transforms are meaningful. | Keep biological checks explicit and domain-specific. |
| "One trait can cover matrices, genomic ranges, variant calls, and single-cell objects." | A huge trait becomes vague and hard to satisfy correctly. | Split contracts by behavior: assay access, interval overlap, variant alleles, sample metadata, and so on. |
| "Runtime traits remove the need for tests." | Runtime conformance only checks method availability or explicit registration. | Use traits as assertions plus ordinary unit tests and validation fixtures. |

A good bioinformatics interface should be boring: small, named after behavior,
well-tested on toy data, and honest about what it leaves to the concrete class.
