---
title: "Bioinformatics Containers as Interfaces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioinformatics Containers as Interfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(S7)
library(s7contract)
```

## Introduction

Bioinformatics packages often exchange rich containers rather than plain
matrices.  Bioconductor's `SummarizedExperiment`, for example, keeps assays,
feature metadata, sample metadata, names, and validity rules synchronized.  This
vignette does not try to rebuild that class.  It uses a small toy container to
show what an interface can and cannot express.

The useful idea is an adapter boundary.  A downstream function might not need a
specific container class.  It might only need to list assay names, retrieve one
assay matrix, and know feature and sample names.  That small behavior can be
written as an S7 interface.

## Background

A class and an interface answer different design questions.  A class describes
representation and invariants: where the assays live, how metadata is stored,
and what must be true after construction or subsetting.  An interface describes
behavior: which operations a consumer may call.

This distinction matters for bioinformatics.  A behavioral interface can make
small examples, tests, and adapters easier to write, but it does not replace a
well-established interoperability class.  It cannot enforce genome builds,
biological interpretation, delayed computation, or the full set of conventions
used by Bioconductor containers.

## A toy assay container

The example class stores a named list of assays plus row and column metadata.
The validator only checks the dimensions needed below.

```{r toy-container}
MiniSummarizedExperiment <- new_class(
  "MiniSummarizedExperiment",
  properties = list(
    assays = class_list,
    row_data = class_data.frame,
    col_data = class_data.frame
  ),
  validator = function(self) {
    if (length(self@assays) == 0) {
      return("@assays must contain at least one matrix")
    }

    dims <- lapply(self@assays, dim)
    if (any(vapply(dims, is.null, logical(1)))) {
      return("every assay must be matrix-like")
    }

    first_dim <- dims[[1]]
    same_dim <- vapply(dims, identical, logical(1), first_dim)
    if (!all(same_dim)) {
      return("all assays must have the same dimensions")
    }

    if (nrow(self@row_data) != first_dim[[1]]) {
      return("@row_data must have one row per assay feature")
    }
    if (nrow(self@col_data) != first_dim[[2]]) {
      return("@col_data must have one row per assay sample")
    }
  }
)

counts <- matrix(
  c(10, 0, 3, 4, 12, 8),
  nrow = 3,
  dimnames = list(c("geneA", "geneB", "geneC"), c("sample1", "sample2"))
)

mini <- MiniSummarizedExperiment(
  assays = list(counts = counts, logcounts = log1p(counts)),
  row_data = data.frame(gc = c(0.42, 0.51, 0.37), row.names = rownames(counts)),
  col_data = data.frame(condition = c("control", "treated"), row.names = colnames(counts))
)
```

## A small assay interface

The interface below describes the behavior needed by a simple consumer.  It is
not a replacement for `SummarizedExperiment`; it is only a view over an object
that has assay-like behavior.

```{r assay-interface}
assay_names <- new_generic("assay_names", "x")
feature_names <- new_generic("feature_names", "x")
sample_names <- new_generic("sample_names", "x")
assay_matrix <- new_generic("assay_matrix", "x")

AssayContainer <- new_interface(
  "AssayContainer",
  methods = list(
    assay_names = assay_names,
    feature_names = feature_names,
    sample_names = sample_names,
    assay_matrix = assay_matrix
  )
)

method(assay_names, MiniSummarizedExperiment) <- function(x) names(x@assays)
method(feature_names, MiniSummarizedExperiment) <- function(x) rownames(x@assays[[1]])
method(sample_names, MiniSummarizedExperiment) <- function(x) colnames(x@assays[[1]])
method(assay_matrix, MiniSummarizedExperiment) <- function(x, name = assay_names(x)[[1]]) {
  x@assays[[name]]
}

implements(mini, AssayContainer)
assay_names(mini)
sample_names(mini)
assay_matrix(mini, "counts")[, "sample1"]
```

A consumer can assert this behavior and then stay independent of the concrete
class.

```{r assay-consumer}
library_size <- function(x, assay = "counts") {
  assert_implements(x, AssayContainer)
  mat <- assay_matrix(x, assay)
  colSums(mat)
}

library_size(mini)
```

This is the productive use case.  A package can write against a small protocol,
while separate adapters provide methods for concrete containers.

## When an explicit trait helps

A trait is useful when structural compatibility is not enough.  Here the trait
records an explicit implementation and stores an associated constant describing
assay orientation.

```{r assay-trait, message = FALSE, warning = FALSE}
ExperimentLike <- new_trait(
  "ExperimentLike",
  methods = list(
    assay_names = trait_method(assay_names),
    feature_names = trait_method(feature_names),
    sample_names = trait_method(sample_names),
    assay_matrix = trait_method(assay_matrix)
  ),
  assoc_consts = c("ASSAY_ORIENTATION")
)

impl_trait(
  ExperimentLike,
  MiniSummarizedExperiment,
  methods = list(
    assay_names = function(x) names(x@assays),
    feature_names = function(x) rownames(x@assays[[1]]),
    sample_names = function(x) colnames(x@assays[[1]]),
    assay_matrix = function(x, name = assay_names(x)[[1]]) x@assays[[name]]
  ),
  assoc_consts = list(ASSAY_ORIENTATION = "features_by_samples"),
  replace = TRUE
)

has_trait(mini, ExperimentLike)
trait_assoc_const(ExperimentLike, mini, "ASSAY_ORIENTATION")
```

## Design cautions

It would be a mistake to define one large trait that tries to cover every
bioinformatics object.  Assay matrices, genomic ranges, variant calls, and
single-cell objects have different invariants and different performance needs.
Small interfaces are easier to satisfy correctly and easier to test.

It would also be a mistake to claim that an interface proves biological
correctness.  Method availability does not prove that samples are comparable,
that row ranges use the same genome build, or that an assay transform is
appropriate for a downstream model.  Those checks should remain explicit and
domain-specific.

The narrow conclusion is useful enough: interfaces can mimic a small behavioral
slice of a class such as `SummarizedExperiment`, but they should not replace the
class or its ecosystem.

## References

- The Bioconductor `SummarizedExperiment` package: <https://bioconductor.org/packages/SummarizedExperiment/>.
- Morgan et al. (2023), "Orchestrating high-throughput genomic analysis with Bioconductor": <https://bioconductor.org/help/publications/>.
- The S7 package documentation: <https://rconsortium.github.io/S7/>.
- The `s7contract` interface and trait vignette in this package.
