---
title: "Contracts Beyond Drawing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contracts Beyond Drawing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

```{r setup}
library(S7)
library(s7contract)
```

The introductory examples for `s7contract` use drawable shapes because the
methods are easy to see.  The same machinery can describe less visual contracts:
"number-like" objects, "vector-like" objects, adapters around storage classes,
or small protocol boundaries between packages.

These examples are intentionally modest.  `s7contract` checks runtime S7 methods;
it does not turn R into a statically typed language.

## A small Number-ish interface

A full mathematical `Number` abstraction is too ambitious for this package:
arithmetic is often binary or multi-dispatch, base vectors have recycling rules,
and numerical code usually needs performance more than abstraction.  A small
single-dispatch interface can still be useful when a function only needs a few
operations.

```{r number-interface}
num_zero <- new_generic("num_zero", "x")
num_add <- new_generic("num_add", "x")
num_scale <- new_generic("num_scale", "x")
num_value <- new_generic("num_value", "x")

NumberLike <- new_interface(
  "NumberLike",
  methods = list(
    zero = num_zero,
    add = num_add,
    scale = num_scale,
    value = num_value
  )
)

method(num_zero, class_double) <- function(x) 0
method(num_add, class_double) <- function(x, y) x + y
method(num_scale, class_double) <- function(x, k) x * k
method(num_value, class_double) <- function(x) x

implements(class_double, NumberLike)
num_add(10, 5)
num_scale(10, 0.5)
```

This is a *protocol* rather than a promise that every numeric law holds.  If you
need associativity, units, precision guarantees, missing-value rules, or
multi-dispatch over both operands, those rules must be documented and tested
separately.

The explicit trait layer can make the contract nominal.  A class has the trait
only after an `impl_trait()` call records it.

```{r number-trait}
Numberish <- new_trait(
  "Numberish",
  methods = list(
    zero = trait_method(num_zero),
    add = trait_method(num_add),
    scale = trait_method(num_scale),
    value = trait_method(num_value)
  ),
  assoc_consts = list(FAMILY = "scalar")
)

impl_trait(
  Numberish,
  class_integer,
  methods = list(
    zero = function(x) 0L,
    add = function(x, y) x + y,
    scale = function(x, k) as.integer(x * k),
    value = function(x) x
  )
)

has_trait(2L, Numberish)
trait_call(Numberish, "add", 2L, 3L)
trait_assoc_const(Numberish, 2L, "FAMILY")
```

## A Vector-ish interface

A vector-like contract can be more practical.  Many algorithms only need to know
how long a thing is, how to slice it, and how to expose its values.

```{r vector-interface}
vec_length <- new_generic("vec_length", "x")
vec_slice <- new_generic("vec_slice", "x")
vec_values <- new_generic("vec_values", "x")

VectorLike <- new_interface(
  "VectorLike",
  methods = list(
    length = vec_length,
    slice = vec_slice,
    values = vec_values
  )
)

ReadDepth <- new_class(
  "ReadDepth",
  properties = list(
    position = class_integer,
    depth = class_double
  ),
  validator = function(self) {
    if (length(self@position) != length(self@depth)) {
      "@position and @depth must have the same length"
    }
  }
)

method(vec_length, ReadDepth) <- function(x) length(x@depth)
method(vec_slice, ReadDepth) <- function(x, i) {
  ReadDepth(position = x@position[i], depth = x@depth[i])
}
method(vec_values, ReadDepth) <- function(x) x@depth

method(vec_length, class_double) <- function(x) length(x)
method(vec_slice, class_double) <- function(x, i) x[i]
method(vec_values, class_double) <- function(x) x

coverage <- ReadDepth(
  position = 1:5,
  depth = c(12, 15, 9, 20, 17)
)

implements(coverage, VectorLike)
vec_length(coverage)
vec_values(vec_slice(coverage, 2:4))
```

A function can assert the small interface it needs and then call only those
required operations.

```{r vector-consumer}
window_mean <- function(x, i) {
  assert_implements(x, VectorLike)
  mean(vec_values(vec_slice(x, i)))
}

window_mean(coverage, 2:4)
window_mean(c(12, 15, 9, 20, 17), 2:4)
```

## Bad ideas to reject early

| Idea | Why it sounds appealing | Critique |
|---|---|---|
| A universal `Number` trait for all arithmetic | One abstraction for integers, doubles, complex numbers, units, sparse arrays, and domain objects | R arithmetic has recycling, missing values, attributes, and binary dispatch needs.  Keep the contract narrow and test algebraic laws separately. |
| A `Vector` trait that replaces base vectors or `vctrs` | One API for every sequence-like object | Existing vector semantics are deep and battle-tested.  Use a tiny protocol at package boundaries, not as a replacement for mature vector frameworks. |
| Traits as static type checks | Compile-time confidence sounds nice | Checks happen at runtime.  They are useful assertions and documentation, not a compiler. |
| Interfaces for every helper function | More contracts can look more rigorous | Too many tiny contracts add ceremony.  Define contracts only at boundaries where substitutability matters. |

The productive pattern is to name a small behavior, state what it does *not*
guarantee, and keep examples small enough to run during `R CMD check`.
