---
title: "A Maybe Dictionary on S7"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Maybe Dictionary on S7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::read_chunk(
  system.file("examples", "maybe-laws.R", package = "s7contract")
)
```

```{r maybe-setup}
```

This S7 dictionary stores `pure` and `bind` as functions. `Just(value)` carries
a value; `Nothing()` represents absence. An interface checks operation
availability, while laws state the required behavior.

## The dictionary

`pure` wraps a value in `Just`. `bind` passes a `Just` payload to a function
returning Maybe, and propagates `Nothing` without calling that function.

```{r maybe-dictionary}
```

Binding `Nothing` leaves the second callback unevaluated:

```{r maybe-use}
```

## Generated values and functions

The input domain is `Nothing` or `Just` containing one integer from -10 to 10,
with equal constructor probabilities. Functions come from three equally likely
families: always return `Nothing`, add an integer from -5 to 5, or retain values
at least a generated threshold in that range. Integer ranges expand with size.
These functions are total on the tested inputs and intermediate values, which
stay between -20 and 20.

```{r maybe-domain}
```

Function descriptions remain data in counterexamples. Shrinking tries earlier
constructors and function families, then moves integer parameters toward zero.
The interpreter constructs each function independently of the dictionary under
test. Equality compares the observable constructor and uses `identical()` for
payloads, so integer and double payloads differ.

## Three laws, one suite

The [Haskell 2010 Report, §6.3.6](https://www.haskell.org/onlinereport/haskell2010/haskellch6.html)
gives the three monad equations. Writing `pure` for the unit operation and
`>>=` for bind:

```text
pure(a) >>= f              = f(a)
m >>= pure                 = m
(m >>= f) >>= g            = m >>= (x -> f(x) >>= g)
```

`maybe_laws(dictionary)` returns three ordinary laws. Each equation uses the
same interpreted functions on both sides. Coverage labels describe reference
outcomes of those functions, including absence at the input or after either
function in a composition.

```{r maybe-law-suite}
```

```{r maybe-checks}
```

The results cover the declared domain; arbitrary R functions, effects, and
payload types are outside it. `dict_bind()` calls the dictionary under test.
`gen_bind()` composes generators and their shrink trees; it does not implement
the dictionary.

## A default that breaks the laws

This dictionary replaces every `Nothing` result with `Just(0L)`. Its operations
still satisfy the structural interface:

```{r maybe-broken}
```

Associativity fails because the default allows a later function to run on one
side of the equation. The reduced input keeps both function descriptions:

```{r maybe-counterexample}
```

The same law and recorded parameters reproduce the failure:

```{r maybe-replay}
```

The vignette and regression tests share
`system.file("examples", "maybe-laws.R", package = "s7contract")`.
The [Haskell monad tutorial](https://www.haskell.org/tutorial/monads.html)
develops the distinction between type-class operations and their laws.
