---
title: "Generative Laws with tinytest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generative Laws with tinytest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(S7)
library(s7contract)
tinytest::using(s7contract)
```

Generative laws separate three concerns: generators construct examples, laws
state behavior over those examples, and a runner searches for a smaller
counterexample. One call to `expect_law()` becomes one tinytest result even
though the law is evaluated many times.

## A tinytest property

The generator below produces integer vectors and carries an integrated shrink
tree. The law checks that reversing a vector twice returns the original value.
This chunk executes while the vignette is built.

```{r reverse-law}
reverse_law <- new_law(
  "reverse is involutive",
  generators = list(
    x = gen_vector(gen_integer(-100L, 100L), max = 20L)
  ),
  holds = function(x) identical(rev(rev(x)), x)
)

expect_law(reverse_law, tests = 100L, seed = 20260902L)
```

## Laws over an S7 interface

A law can exercise an existing `s7contract` interface. The generator constructs
valid `Circle` objects rather than attempting to invert the class validator.
Calls inside the law retain the interface's return contract.

```{r interface-law}
Circle <- new_class(
  "CirclePropertyVignette",
  properties = list(radius = class_double),
  validator = function(self) {
    if (self@radius < 0) "`radius` must be non-negative."
  }
)
area <- new_generic(
  "area_property_vignette",
  "x",
  function(x) S7_dispatch()
)
method(area, Circle) <- function(x) pi * x@radius^2

HasArea <- new_interface(
  "HasAreaPropertyVignette",
  generics = list(
    area = interface_requirement(area, returns = class_double)
  )
)
circles <- gen_map(
  gen_integer(0L, 1000L),
  function(radius) Circle(radius = as.double(radius))
)
area_law <- new_law(
  "non-negative radii have non-negative area",
  generators = list(x = circles),
  holds = function(x) with(HasArea, area(x)) >= 0
)

expect_law(area_law, tests = 100L, seed = 20260902L)
```

## Inspecting and replaying a failure

Use `check_law()` when the structured result is needed independently of a test
framework. This deliberately false law starts at ten and shrinks to zero.

```{r counterexample}
ten <- new_generator(
  draw = function(size) 10L,
  shrink = function(value) {
    if (value == 0L) list() else list(0L, value %/% 2L)
  },
  label = "ten",
  prototype = integer()
)
negative_law <- new_law(
  "generated values are negative",
  generators = list(x = ten),
  holds = function(x) x < 0L
)

failure <- check_law(negative_law, tests = 10L, seed = 20260902L)
failure
```

The result records the seed, RNG kind, run parameters, original input, smallest
counterexample found, and shrink counts. The `minimal` field keeps its name for
compatibility; it does not promise a global minimum. `shrink_status` distinguishes
completion within the shrink tree, an evaluation budget, a shrinking error, and
a run that needed no shrinking. If shrinking errors or warns, `shrink_condition`
records that problem while the original and last failing examples remain available.

Replay the same law and parameters with ordinary R function application:

```{r replay}
replayed <- do.call(check_law, c(list(law = failure@law), failure@parameters))
identical(replayed@counterexample@minimal, failure@counterexample@minimal)
```

Runs use Mersenne-Twister, Inversion normals, and Rejection sampling, so changing
the caller's RNG kind does not change the generated sequence. Replay requires
unchanged generator and law code, run parameters, and compatible R/package
versions. Generators and laws must not depend on external mutable state or
change the RNG configuration. Stored examples can still be tested directly when
the generator changes.

The caller's RNG kind and state are restored on exit. The exception is an
admission restriction: callers using Box-Muller normals are rejected before
anything changes, because R does not expose their cached normal draw for
restoration. Select another normal RNG kind before running laws in that session.

## Composition and shrinking

Mapping transforms both the generated value and visited shrinks. Products
combine independent generators and shrink one component at a time. Vectors
remove contiguous chunks and then shrink elements, retaining their minimum
length. Nesting vector generators produces lists of vectors, including empty
inner vectors:

```{r nested-vectors}
nested <- new_law(
  "nested vectors retain their element type",
  generators = list(x = gen_vector(gen_vector(gen_integer(), max = 4L), max = 3L)),
  holds = function(x) is.list(x) && all(vapply(x, is.integer, logical(1)))
)
expect_law(nested, tests = 20L, seed = 1L)
```

The runner constructs and transforms each shrink candidate only when visited.
`shrinks = 0L` performs no shrink expansion. A custom `shrink` function still
constructs its own list of candidates; the evaluation budget cannot bound the
work performed inside user functions. Shrinkers and mapping functions must be
deterministic, and mapped constructors must accept every visited shrink.

## Relationship to Hedgehog

[R Hedgehog](https://hedgehogqa.r-universe.dev/hedgehog) is the reference for the
broader property-testing scope. Its generators carry lazy rose trees, with
deterministic shrinking preserved through composition. The underlying
[Haskell Hedgehog](https://hackage.haskell.org/package/hedgehog) design includes
mapping, dependent generation, size-aware ranges, and state-machine testing.

The table describes the scope of s7contract 0.2.0. It is not a compatibility claim.

| Concept | s7contract 0.2.0 |
|:--|:--|
| Mapping / functor composition | `gen_map()` transforms values and their shrink trees. |
| Independent / applicative composition | `gen_product()` and named law arguments combine independent generators. |
| Dependent / monadic composition | No built-in bind combinator; custom generators must supply generation and shrinking together. |
| Size-aware generation | Integer ranges and vector lengths grow with the runner's size, up to configured bounds. |
| Choice and recursive generation | No built-in combinators. |
| State-machine testing | No command/model runner. |
| Behavioral contracts | Laws can exercise S7 interfaces and traits through ordinary calls. |
| Test-framework integration | `check_law()` returns structured results; `expect_law()` records one tinytest result. |
