---
title: "Generative Laws with tinytest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generative Laws with tinytest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(S7)
library(s7contract)
tinytest::using(s7contract)
```

Generative laws separate three concerns: generators construct examples, laws
state behavior over those examples, and a runner searches for a smaller
counterexample. One call to `expect_law()` becomes one tinytest result even
though the law is evaluated many times.

In `s7contract`, these laws provide behavioral evidence for protocols whose
operations are described by interfaces or traits. The
[vector protocol](protocol-laws.html) vignette defines one
`VectorLike` law suite, runs it against two implementations, and finds a faulty
slice method that satisfies the interface.

## A tinytest property

The generator below produces integer vectors and carries an integrated shrink
tree. The law checks that reversing a vector twice returns the original value.

```{r reverse-law}
reverse_law <- new_law(
  "reverse is involutive",
  generators = list(
    x = gen_vector(gen_integer(-100L, 100L), max = 20L)
  ),
  holds = function(x) identical(rev(rev(x)), x)
)

expect_law(reverse_law, tests = 100L, seed = 20260902L)
```

## Laws over an S7 interface

Generate non-negative radii to construct valid `Circle` objects. The law checks
their areas through an interface that requires a double return value.

```{r interface-law}
Circle <- new_class(
  "CirclePropertyVignette",
  properties = list(radius = class_double),
  validator = function(self) {
    if (self@radius < 0) "`radius` must be non-negative."
  }
)
area <- new_generic(
  "area_property_vignette",
  "x",
  function(x) S7_dispatch()
)
method(area, Circle) <- function(x) pi * x@radius^2

HasArea <- new_interface(
  "HasAreaPropertyVignette",
  generics = list(
    area = interface_requirement(area, returns = class_double)
  )
)
circles <- gen_map(
  gen_double(0, 1000),
  function(radius) Circle(radius = radius)
)
area_law <- new_law(
  "non-negative radii have non-negative area",
  generators = list(x = circles),
  holds = function(x) with(HasArea, area(x)) >= 0
)

expect_law(area_law, tests = 100L, seed = 20260902L)
```

## Inspecting and replaying a failure

Use `check_law()` when the structured result is needed independently of a test
framework. This deliberately false law starts at ten and shrinks to zero.

```{r counterexample}
ten <- new_generator(
  draw = function(size) 10L,
  shrink = function(value) {
    if (value == 0L) list() else list(0L, value %/% 2L)
  },
  label = "ten",
  prototype = integer()
)
negative_law <- new_law(
  "generated values are negative",
  generators = list(x = ten),
  holds = function(x) x < 0L
)

failure <- check_law(negative_law, tests = 10L, seed = 20260902L)
failure
```

The result records the seed, RNG kind, run parameters, original input, smallest
counterexample found, and shrink counts. The `minimal` field holds the smallest
failure found within the shrink tree and evaluation budget.
`shrink_status` distinguishes completion within the shrink tree, an evaluation
budget, a shrinking error, and a run that needed no shrinking. If shrinking
errors or warns, `shrink_condition` records that problem while the original and
last failing examples remain available.

Replay the same law and parameters with ordinary R function application:

```{r replay}
replayed <- do.call(check_law, c(list(law = failure@law), failure@parameters))
identical(replayed@counterexample@minimal, failure@counterexample@minimal)
```

Runs use Mersenne-Twister, Inversion normals, and Rejection sampling, so changing
the caller's RNG kind does not change the generated sequence. Replay requires
unchanged generator and law code, run parameters, and compatible R/package
versions. Generators and laws must not depend on external mutable state or
change the RNG configuration. Stored examples can still be tested directly when
the generator changes.

The caller's RNG kind and state are restored on exit. Box-Muller normals are
unsupported because R does not expose their cached draw for restoration; select
another normal RNG kind before running laws.

## Composition and shrinking

Mapping transforms both the generated value and visited shrinks. Products
combine independent generators and shrink one component at a time. Vectors
remove contiguous chunks and then shrink elements, retaining their minimum
length. Nesting vector generators produces lists of vectors, including empty
inner vectors:

```{r nested-vectors}
nested <- new_law(
  "nested vectors retain their element type",
  generators = list(x = gen_vector(gen_vector(gen_integer(), max = 4L), max = 3L)),
  holds = function(x) is.list(x) && all(vapply(x, is.integer, logical(1)))
)
expect_law(nested, tests = 20L, seed = 1L)
```

The runner constructs and transforms each shrink candidate only when visited.
`shrinks = 0L` performs no shrink expansion. A custom `shrink` function still
constructs its own list of candidates; the evaluation budget cannot bound the
work performed inside user functions. Shrinkers and mapping functions must be
deterministic, and mapped constructors must accept every visited shrink.

## Dependent inputs

`gen_bind()` uses one generated value to construct the next generator. Here the
sequence length and its bases belong to one dependent input. Every shrink still
has exactly the declared length, so the law needs no discarded preconditions.

```{r dependent-sequences}
sequences <- gen_bind(gen_integer(0L, 20L), function(n) {
  gen_product(
    length = gen_constant(n),
    bases = gen_vector(gen_element(c("A", "C", "G", "T")), min = n, max = n)
  )
})
sequence_law <- new_law(
  "sequence length matches its declaration",
  generators = list(x = sequences),
  holds = function(x) length(x$bases) == x$length
)
expect_law(sequence_law, tests = 40L, seed = 1L)
gen_example(sequences, size = 10L, seed = 42L)
```

Shrinking first tries smaller source values and rebuilds the dependent
generator, then shrinks its result. Each rebuild uses the same captured local
seed and size. Random draws inside the law therefore do not change the
regenerated candidates. The factory must depend only on its input and the
scoped RNG, and its constructors must accept every visited source shrink.

## Choices and recursive values

`gen_element()` chooses a value; `gen_choice()` chooses a generator. Entries are
ordered from simpler to more complex for shrinking. Optional `prob` weights
control sampling; zero-weight entries are excluded from generation and
shrinking. All entries with positive weight are available even at size zero.

```{r nullable-values}
nullable <- gen_choice(gen_constant(NA_integer_), gen_integer(), prob = c(1, 9))
gen_example(gen_vector(nullable, min = 6L, max = 6L), size = 10L, seed = 42L)
```

`gen_sized()` builds a generator from the runner's current size. `gen_resize()`
overrides the size for one generator while leaving sibling generators alone.
`gen_recursive()` supplies its expansion function with a child generator that
uses half the current size, rounded down. At size zero, only the base generator
runs. This supports nested lists, expression trees, or recursive S7 objects.

```{r recursive-values}
trees <- gen_recursive(
  gen_element(c("A", "C", "G", "T")),
  function(child) gen_product(left = child, right = child)
)
gen_example(trees, size = 7L, seed = 42L)

leaf_count <- function(tree) {
  if (is.list(tree)) leaf_count(tree$left) + leaf_count(tree$right) else 1L
}
tree_law <- new_law(
  "binary trees have at least one leaf",
  generators = list(tree = trees),
  holds = function(tree) leaf_count(tree) >= 1L
)
expect_law(tree_law, tests = 30L, seed = 1L, max_size = 7L)
```

Recursive size bounds depth, not total node count; the expansion function still
controls branching. Shrinking can replace a recursive value with a base value
before shrinking within a branch. `gen_no_shrink()` removes a generator's
shrinking when a value must remain fixed during the search. `gen_example()`
draws one value without expanding shrinks and restores the caller's RNG state.

## Relationship to Hedgehog

Generation and shrinking follow
[R Hedgehog](https://hedgehogqa.r-universe.dev/hedgehog) and
[Haskell Hedgehog](https://hackage.haskell.org/package/hedgehog): generators
carry lazy rose trees, preserving shrinking through composition. The
corresponding operations in `s7contract` are:

| Concept | s7contract |
|:--|:--|
| Mapping / functor composition | `gen_map()` transforms values and their shrink trees. |
| Independent / applicative composition | `gen_product()` and named law arguments combine independent generators. |
| Dependent / monadic composition | `gen_bind()` rebuilds downstream generators when upstream inputs shrink. |
| Size-aware generation | Integer ranges and vector lengths grow with size; `gen_sized()` and `gen_resize()` expose size control. |
| Choice and recursive generation | `gen_element()`, `gen_choice()`, and `gen_recursive()` retain integrated shrinking. |
| Inspection and shrink control | `gen_example()` draws a reproducible value; `gen_no_shrink()` removes shrinking. |
| State-machine testing | `new_command()`, `gen_commands()`, and `new_state_law()` test sequential protocols against a model. |
| Case coverage | `classify` labels generated inputs; `min_coverage` requires observed proportions within the test budget. See the [vector example](protocol-laws.html#which-cases-were-tested). |
| Numeric domains | `gen_double()` generates finite fractional values with a shrink origin; `gen_choice()` adds exceptional values with explicit weights. |
| Selection domains | `gen_sample()` preserves sample cardinality and distinct source positions; `gen_subsequence()` preserves source order. See the [vector laws](protocol-laws.html). |
| Strings and calendar dates | Composed recipes exercise [UTF-8 store keys](stateful-protocols.html) and [whole-day intervals](calendar-intervals.html). |
| Behavioral contracts | Laws can exercise S7 interfaces and traits through ordinary calls. |
| Test-framework integration | `check_law()` returns structured results; `expect_law()` records one tinytest result. |
