---
title: "Generative Laws with tinytest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generative Laws with tinytest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(S7)
library(s7contract)
tinytest::using(s7contract)
```

Generative laws separate three concerns: generators construct examples, laws
state behavior over those examples, and a runner searches for a minimized
counterexample. One call to `expect_law()` becomes one tinytest result even
though the law is evaluated many times.

## A tinytest property

The generator below produces integer vectors and carries an integrated shrink
tree. The law checks that reversing a vector twice returns the original value.
This chunk executes while the vignette is built.

```{r reverse-law}
reverse_law <- new_law(
  "reverse is involutive",
  generators = list(
    x = gen_vector(gen_integer(-100L, 100L), max = 20L)
  ),
  holds = function(x) identical(rev(rev(x)), x)
)

expect_law(reverse_law, tests = 100L, seed = 20260902L)
```

## Laws over an S7 interface

A law can exercise an existing `s7contract` interface. The generator constructs
valid `Circle` objects rather than attempting to invert the class validator.
Calls inside the law retain the interface's return contract.

```{r interface-law}
Circle <- new_class(
  "CirclePropertyVignette",
  properties = list(radius = class_double),
  validator = function(self) {
    if (self@radius < 0) "`radius` must be non-negative."
  }
)
area <- new_generic(
  "area_property_vignette",
  "x",
  function(x) S7_dispatch()
)
method(area, Circle) <- function(x) pi * x@radius^2

HasArea <- new_interface(
  "HasAreaPropertyVignette",
  generics = list(
    area = interface_requirement(area, returns = class_double)
  )
)
circles <- gen_map(
  gen_integer(0L, 1000L),
  function(radius) Circle(radius = as.double(radius))
)
area_law <- new_law(
  "non-negative radii have non-negative area",
  generators = list(x = circles),
  holds = function(x) with(HasArea, area(x)) >= 0
)

expect_law(area_law, tests = 100L, seed = 20260902L)
```

## Inspecting a minimized failure

Use `check_law()` when the structured result is needed independently of a test
framework. This deliberately false law starts at ten and shrinks to zero.

```{r counterexample}
ten <- new_generator(
  draw = function(size) 10L,
  shrink = function(value) {
    if (value == 0L) list() else list(0L, value %/% 2L)
  },
  label = "ten",
  prototype = integer()
)
negative_law <- new_law(
  "generated values are negative",
  generators = list(x = ten),
  holds = function(x) x < 0L
)

check_law(negative_law, tests = 10L, seed = 20260902L)
```

The result records the seed, original input, minimized counterexample, and
shrink counts. Seed replay remains deterministic for an unchanged generator;
the stored minimized value remains useful if generator code later changes.
