---
title: "Reusable Laws for a Vector Protocol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reusable Laws for a Vector Protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::read_chunk(
  system.file("examples", "vector-laws.R", package = "s7contract")
)
```

```{r setup}
library(S7)
library(s7contract)
```

This example extends [Behavioral Contracts on S7](s7-interfaces-and-traits.html)
with one law suite shared by two implementations and a faulty subclass.

## Vector-like behavior

Many algorithms need only a length, a way to slice, and access to values. Both
ordinary double vectors and `ReadDepth` objects implement those operations.
The class validator keeps positions and depths aligned, while the interface
describes the behavior consumers need.

```{r vector-interface}
```

A function can depend on this small protocol without knowing how the object is
represented internally.

```{r vector-consumer}
```

## One protocol, several implementations

A protocol author can publish a function returning a named list of laws.
Implementation authors supply a constructor; the laws compare each result with
the reference values passed to that constructor.

The domain here is unnamed double vectors containing small integers, with
positive, in-range integer indices. Empty vectors and selections are included;
indices may repeat or appear out of order. Missing values, names, negative
indices, and the rest of R's subsetting semantics are outside this example.

`gen_bind()` constructs the object and an index generator from the reference
values. When those values shrink, it rebuilds both, preserving object validity
and index bounds. Bounds come from the reference data rather than the method
being tested. Inside the contract mask, `base::length()` keeps the reference
calculation separate from the interface's `length` alias.

```{r vector-law-suite}
```

`lapply()` runs the same four laws against both representations.

```{r vector-law-implementations}
```

## Which cases were tested?

The slicing law classifies inputs as empty or nonempty, and labels selections
with repeated or reordered indices. Each label counts once per accepted case.
Its `min_coverage` requirements use proportions: `reordered = 0.1` asks for
reordered indices in at least 10% of cases.

```{r vector-law-coverage}
```

Fixing size at zero exercises only empty vectors. The predicate passes, but the
coverage requirements prevent the run from passing:

```{r vector-law-insufficient-coverage}
```

This follows the test-data classification discussed by
[Claessen and Hughes (2000), §2.4](https://users.cs.northwestern.edu/~robby/courses/395-495-2009-fall/quick.pdf).
These minima describe observed proportions within the chosen test budget;
they carry no statistical confidence guarantee. Discards, errors, and shrink
evaluations contribute no counts. A falsifying generated case does count, and
a run that ends early reports partial coverage alongside its primary failure.
See `new_law()` for classifier requirements and result fields.

## Structural conformance and a behavioral failure

This subclass inherits the correct length and value methods, but its slice
method reverses the requested order. All required methods are available, so
`implements()` succeeds. Its slices also have valid representations and the
expected value type and length.

```{r vector-law-broken}
```

Only the law about selected values and their order fails. `check_law()` reports
that failure separately from structural conformance, and shrinks it to a smaller
valid input. The checked call below still succeeds; its result differs from the
reference slice.

```{r vector-law-counterexample}
```

The result records the inputs and run parameters needed to replay the failure:

```{r vector-law-replay}
```

These example definitions and checks are installed together in
`system.file("examples", "vector-laws.R", package = "s7contract")`. The vignette
and package tests execute that same script. For generator composition, budgets,
and tinytest integration, see `vignette("property-laws")`.

`implements()` continues to check method availability; `has_trait()` checks
declared implementation. Neither runs laws or changes meaning after a law passes
or fails. The protocol's laws and its implementation-specific generators are
explicit test inputs.
